---
title: "Distribution for Quadratic Form Estimators"
date: "2/27/2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyr)
library(ggplot2)
library(dplyr)
library(MASS) # Required for mvrnorm

mtse=modules::use("C:/Users/aak3/Documents/atomic-clock/Functions.R")
source("C:/Users/aak3/Documents/atomic-clock/Functions_SpectrumCovariance.R")

```

Let $Y = \mathbf{X}^{T} \mathbf{A} \mathbf{X}$ be a quadratic form, where $\mathbf{A}$ is a real $n \times n$ symmetric matrix and $\mathbf{X} \sim N_{n}(\boldsymbol{\mu}, \Sigma)$ with $\Sigma > 0$.

Spectral Decomposition: $\Sigma = \Sigma^{1/2}\Sigma^{1/2}$, where $\Sigma^{1/2}$ is a positive definite $(\Sigma^{1/2} > 0)$ symmetric matrix. Then:

$$\Sigma^{-1/2}\mathbf{X} \sim N_{n}(\Sigma^{-1/2}\boldsymbol{\mu}, I)$$

The quadratic form $Y$ can be rewritten by inserting the identity matrix and applying spectral decomposition:

$$Y = \mathbf{X}^{T} \mathbf{A} \mathbf{X} = \mathbf{X}^{T}\Sigma^{-1/2}\Sigma^{1/2}\mathbf{A}\Sigma^{1/2}\Sigma^{-1/2}\mathbf{X} = (\Sigma^{-1/2}\mathbf{X})^{T}P\Lambda P^{T}(\Sigma^{-1/2}\mathbf{X})$$

* Here, $P\Lambda P^{T}$ represents the spectral decomposition of $\Sigma^{1/2}A\Sigma^{1/2}$.
* $\Lambda$ is a diagonal matrix of eigenvalues $\lambda_{i}$ and $P$ is orthogonal

So $Y = \mathbf{W}^{T}\Lambda\mathbf{W}$ where
$$\mathbf{W} = P^{T}\Sigma^{-1/2}\mathbf{X}$$

* $\mathbf{W} \sim N_{n}(\boldsymbol{\nu}, I)$ 
* $\boldsymbol{\nu} = P^{T}\Sigma^{-1/2}\boldsymbol{\mu}$


Finally, $$Y = \sum_{i=1}^{r} \lambda_{i}W_{i}^{2} \quad \text{where} \quad W_{i}^{2} \sim \chi^{2}(1, \nu_{i}^{2})$$


* In other words, $Y$ is a weighted sum of $r$ independent non-central chi-square random variables, each with 1 degree of freedom.
* $r$ is the rank of matrix $A$ ($r = \text{rank}(A)$).
* The terms $\nu_{i}$ are the components of the vector $\boldsymbol{\nu} = P^{T}\Sigma^{-1/2}\boldsymbol{\mu}$.


## Simplification

For $\boldsymbol{\mu}=\mathbf{0}$, as in our case

$$Y = \sum_{i=1}^{r} \lambda_{i}W_{i}^{2} \quad \text{where} \quad W_{i}^{2} \sim \chi^{2}(1)$$


* $Y$ is a weighted sum of $r$ independent central chi-square random variables, each with 1 degree of freedom.
* $r = \text{rank}(A)$ = Number of tapers for MT spectral estimate

## Distributional approximation

To approximate the distribution of the spectral quadratic form, we use a two-moment match to a scaled chi-squared distribution (Satterthwaite, 1946; Box, 1954). These papers show that you can approximate the distribution of the weighted sum of chi-square random variables with 1 degree of freedom, $Y = \sum_{i=1}^{r} \lambda_{i}W_{i}^{2}$, as 
$$ \alpha \chi^{2}(\nu_{\text{eff}}) $$
where 
$$ \alpha = \frac{\sum_{i=1}^{r} \lambda_{i}^2}{\sum_{i=1}^{r} \lambda_{i}}$$
and the effective degrees of freedom 
$$ \nu_{\text{eff}} = \frac{(\sum_{i=1}^{r} \lambda_{i})^2}{\sum_{i=1}^{r} \lambda_{i}^2}$$

## Code example 

```{r}

#################################################################
# Quadratic Form Distribution via Weighted Chi-Squares
#################################################################

######################################
# 1. Setup Parameters
######################################

N <- 100  # length of the time series
ar_coeffs <- c(0.5, -0.4, 0.3, -0.2)  # AR(4) coefficients

mu=rep(0,N)

t.vec <- 1:N

# # Simulate AR(4) process
x.t <- arima.sim(model = list(ar = ar_coeffs), n = N)
x.t=x.t-mean(x.t) #we always demean

max.lag.acf=9
s_acf <- stats::ARMAacf(ar = ar_coeffs, ma = numeric(0), lag.max = max.lag.acf)

# Create a Toeplitz matrix from the autocorrelation values
N.noNA=sum(!is.na(t.vec))

R_mat <- matrix(0, nrow = N.noNA, ncol = N.noNA)
R_mat <- stats::toeplitz(c(s_acf, rep(0, N.noNA - max.lag.acf - 1))) 

Sigma <- R_mat

# Matrix A (Sample Variance Kernel)
I <- diag(N)
J <- matrix(1, N, N)

myK=10
W1factor=12
V.mat3 <- mtse$get_tapers(t.vec, W = W1factor/N, K = myK)
taperMat=V.mat3$tapers
V.mat3$e.values

N.fourier=N/2+1

freq <- seq(0, 0.5, length.out = N.fourier)

#### do for one frequency
i=2
U <- taperMat*exp(-1i*2*pi*freq[i]*t.vec)/sqrt(myK)
U_star <- Conj(t(U))
A_f <- U %*% U_star

sum(Im(A_f)) #hopefully all 0 so i can just take the real part below

A<-Re(A_f)

# --- 2. Eigen-Decomposition for Weights ---
# The weights (lambdas) of the quadratic form are the eigenvalues of A %*% Sigma
# We use the symmetric version: Sigma^(1/2) %*% A %*% Sigma^(1/2)
eig_Sig <- eigen(Sigma)
Sigma_half <- eig_Sig$vectors %*% diag(sqrt(eig_Sig$values)) %*% t(eig_Sig$vectors)

M <- Sigma_half %*% A %*% Sigma_half
lambdas <- eigen(M)$values
lambdas <- lambdas[lambdas > 1e-10] # Filter numerical noise

# --- 3. Calculate Approximation Parameters ---
s1 <- sum(lambdas)
s2 <- sum(lambdas^2)
s3 <- sum(lambdas^3)

# Satterthwaite (2-moment Gamma)
alpha_s <- s2 / s1
nu_s    <- s1^2 / s2

# Another alternative: Liu-Tang-Zhang (3-moment Shifted Gamma)
alpha_ltz <- s3 / s2
nu_ltz    <- s2^3 / s3^2
delta     <- s1 - (s2^2 / s3)

# --- 4. Simulation Comparison ---
n_sim <- 10000

# A. Ground Truth: Direct Quadratic Form
sim_direct <- replicate(n_sim, {
  x <- mvrnorm(1, mu, Sigma)
  as.numeric(t(x) %*% A %*% x)
})

# B. Theory: Weighted Sum of Independent Chi-Squares
# (Since mu=0, NCP is 0)
sim_weighted <- matrix(rchisq(n_sim * length(lambdas), 1), ncol=length(lambdas)) %*% lambdas

# C. Satterthwaite Approximation (Scaled Chi-Square)
sim_satt <- alpha_s * rchisq(n_sim, df = nu_s)

# D. Liu-Tang-Zhang (Shifted & Scaled Chi-Square)
sim_ltz  <- alpha_ltz * rchisq(n_sim, df = nu_ltz) + delta

# --- 5. Final Visualization ---
simres <- data.frame(
  values = c(as.vector(sim_direct), as.vector(sim_weighted), 
             as.vector(sim_satt), as.vector(sim_ltz)),
  type = rep(c("1. Direct (x'Ax)", "2. Weighted Sum", 
               "3. Satterthwaite", "4. Liu-Tang-Zhang"), each = n_sim)
)

ggplot(simres, aes(x = values, color = type)) +
  geom_density(linewidth = 1) +
  labs(title = "Quadratic Form Distribution Accuracy",
       subtitle = "Comparing Direct Simulation vs. Weighted Sum vs. Gamma Approximations",
       x = "Value", y = "Density", color = "Method") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

# Numerical check of the means

```

```{r,echo=FALSE}
cat("Theoretical Mean:  ", s1, "\n")
cat("Direct Mean:       ", mean(sim_direct), "\n")
cat("Satterthwaite Mean:", mean(sim_satt), "\n")
cat("LTZ Mean:          ", mean(sim_ltz), "\n")

```


## To do

Above shows that this works for a basic example. For larger N, what kind of matrix calculations do I really have to do to get this approximate distribution?

## References

Satterthwaite, F. E. (1946). An Approximate Distribution of Estimates of Variance Components. Biometrics Bulletin, 2(6), 110–114.

Box, G. E. P. (1954). Some theorems on quadratic forms applied in the study of analysis of variance problems, I. Effect of inequality of variance in the one-way classification. The Annals of Mathematical Statistics, 25(2), 290–302.




## Extend to many frequencies

```{r}

######################################
# 1. Setup Parameters
######################################

N <- 100  # length of the time series
ar_coeffs <- c(0.5, -0.4, 0.3, -0.2)  # AR(4) coefficients

mu=rep(0,N)

t.vec <- 1:N

# # Simulate AR(4) process
x.t <- arima.sim(model = list(ar = ar_coeffs), n = N)
x.t=x.t-mean(x.t) #we always demean

max.lag.acf=9
s_acf <- stats::ARMAacf(ar = ar_coeffs, ma = numeric(0), lag.max = max.lag.acf)

# Create a Toeplitz matrix from the autocorrelation values
N.noNA=sum(!is.na(t.vec))

R_mat <- matrix(0, nrow = N.noNA, ncol = N.noNA)
R_mat <- stats::toeplitz(c(s_acf, rep(0, N.noNA - max.lag.acf - 1))) 

Sigma <- R_mat

# Matrix A (Sample Variance Kernel)
I <- diag(N)
J <- matrix(1, N, N)

myK=10
W1factor=12
V.mat3 <- mtse$get_tapers(t.vec, W = W1factor/N, K = myK)
taperMat=V.mat3$tapers
V.mat3$e.values

N.fourier=N/2+1

freq <- seq(0, 0.5, length.out = N.fourier)

#### do for one frequency
i=2
U <- taperMat*exp(-1i*2*pi*freq[i]*t.vec)/sqrt(myK)
U_star <- Conj(t(U))
A_f <- U %*% U_star

sum(Im(A_f)) #hopefully all 0 so i can just take the real part below

A<-Re(A_f)

# --- 2. Eigen-Decomposition for Weights ---
# The weights (lambdas) of the quadratic form are the eigenvalues of A %*% Sigma
# We use the symmetric version: Sigma^(1/2) %*% A %*% Sigma^(1/2)
eig_Sig <- eigen(Sigma)
Sigma_half <- eig_Sig$vectors %*% diag(sqrt(eig_Sig$values)) %*% t(eig_Sig$vectors)

M <- Sigma_half %*% A %*% Sigma_half
lambdas <- eigen(M)$values
lambdas <- lambdas[lambdas > 1e-10] # Filter numerical noise

# --- 3. Calculate Approximation Parameters ---
s1 <- sum(lambdas)
s2 <- sum(lambdas^2)
s3 <- sum(lambdas^3)

# Satterthwaite (2-moment Gamma)
alpha_s <- s2 / s1
nu_s    <- s1^2 / s2

# Another alternative: Liu-Tang-Zhang (3-moment Shifted Gamma)
alpha_ltz <- s3 / s2
nu_ltz    <- s2^3 / s3^2
delta     <- s1 - (s2^2 / s3)

# B. Theory: Weighted Sum of Independent Chi-Squares
# (Since mu=0, NCP is 0)
sim_weighted <- matrix(rchisq(n_sim * length(lambdas), 1), ncol=length(lambdas)) %*% lambdas

# C. Satterthwaite Approximation (Scaled Chi-Square)
sim_satt <- alpha_s * rchisq(n_sim, df = nu_s)

# D. Liu-Tang-Zhang (Shifted & Scaled Chi-Square)
sim_ltz  <- alpha_ltz * rchisq(n_sim, df = nu_ltz) + delta

# --- 5. Final Visualization ---
simres <- data.frame(
  values = c(as.vector(sim_weighted), 
             as.vector(sim_satt), as.vector(sim_ltz)),
  type = rep(c("Weighted Sum", 
               "Satterthwaite", "Liu-Tang-Zhang"), each = n_sim)
)

ggplot(simres, aes(x = values, color = type)) +
  geom_density(linewidth = 1) +
  labs(x = "Value", y = "Density", color = "Method") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")


```



## Next direction

What is the distribution of:

\begin{equation}
    \widehat{\sigma}^2_{\text{SPEC}}(\tau) = \sum_j \Delta_f G_\tau(f_j) \widehat{S}(f_j)=\Delta_f \bG_\tau^T \widehat{\bS},
\end{equation}

Is a weighted sum of correlated, scaled, shifted chi squared (approximated!) also a scaled shifted chi squared that I can MoM match to approximate? 